---
title: "POLSC 343 US Foreign Policy"
date: "`r Sys.Date()`"
bibliography: "bib/classrefs-polsc343-2021.bib"
csl: "bib/apsr-bibliography.csl"
output: 
  distill::distill_article:
    toc: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

mon <- as.Date("2020-08-23")

advdate <- function(obj, adv){
  tmon <- obj + 7*(adv - 1)
  tfri <- obj + 4 + 7*(adv - 1)
  tmon <- format(tmon, format = "%m/%d")
  tfri <- format(tfri, format = "%m/%d")
  zadv <- sprintf("%02d", adv)
  tmp <- paste("Week ", zadv, sep = '', ", ", tmon, " - ", tfri)
  return(tmp)
}

```
 

## Course Summary
 
This course focuses on the development of US foreign policy over the
course of the twentieth and twenty-first centuries. It will provide a brief background
on the history of US foreign policy before 1900, but will focus primarily on the developments and issues
occurring from 1900 through the present. The historical content of the course will be supplemented by
introducing students to political science literature that can help to identify and explain key events and trends
in US foreign policy during this time period. 

## Course Objectives

1. The overarching goal is to develop an understanding
of the systematic forces that shape US foreign policymaking. Please note that this is different than
prescribing what policy *should* be (though this is also an important consideration when considering our motivations for studying the topic and reading course materials.
2. Identify the key actors that influence foreign policy.
3. Understand how and when the influence of actors varies (e.g. across time, issue area, etc.).
4. Identify the key issues confronting policymakers and society. 
5. Identify general trends in US foreign policy over time. 
    - What's changed? 
    - What stays the same?
    - When and why do we see changes in behavior or policies?
6. Students should develop a familiarity with social science research processes, practices. Specifically, students should have:
    - A basic understanding of how to read social scientific research articles. 
    - A basic understanding of research design considerations
    - A basic understanding of what the scope and limitations of research articles


## Course Format 

The course will be part lecture and part student participation. I encourage students to ask questions if the readings of lecture materials are unclear. In addition to class lectures students will be responsible for keeping up with readings and other class-related material as the semester progresses.


## Covid-19 Planning and Contingencies 

As you may have noticed, we're still in the middle of a pandemic. The Delta variant spreads much easier and more quickly than earlier variants of the virus and poses a real risk to our ability to continue in-person instruction this year. We have three primary tools to help us continue in-person instruction, because, frankly, online learning isn't fun for anyone:

1. Vaccination
2. Masking
3. Ventilation 

**Vaccines**: Vaccination remains our best option to help prevent the spread of, and mitigate the worst effects of, COVID-19. I will not ask your vaccination status at any point, but I strongly urge students who are not currently vaccinated and who are able to talk to your physicians about getting vaccinated. It's safe, effective, and free. You can find information and/or make an appointment at [Lafene Health Center](https://www.k-state.edu/covid-19/vaccine/), [The Riley County Health Department](https://www.rileycountyks.gov/1916/Vaccine-Information), local pharmacies, or your primary care physician.  

**Masks**: Please note that K-State currently has a mask requirement in effect. That means that all students, faculty, staff, and visitors are required to mask while inside buildings at K-State. To state the obvious, this includes students attending lectures. Also please note that some masks are better than others---N95, KN95, FFP2, and similar types of masks offer strong protection against transmission (for example, see [here](https://www.sciencedirect.com/science/article/pii/S0196655320310439?casa_token=3QaMU4BJHh0AAAAA:luVTg-SgOeHvQl3F7_vj54R8RvbEWIwVjLLbs4L6JG7QkbCHfYVRPRiQmfHcRzZDfzOcm5-br4o) and [here](https://www.tandfonline.com/doi/full/10.1080/02786826.2020.1862409)). Gaiters also don't offer as much protection as higher quality masks. Face shields do effectively nothing on their own. You can find more information [here on Emergency Use Authorizations](https://www.fda.gov/medical-devices/coronavirus-disease-2019-covid-19-emergency-use-authorizations-medical-devices/personal-protective-equipment-euas) for various mask types,

**Ventilation**: In addition to personal preventative measures, ventilation and air filtration are enormously important in helping to reduce individuals' exposure to the virus. K-State has conducted a review of campus rooms over the last year to determine which rooms needed upgrades to their ventilation systems to ensure proper air flow. 

---

**A Personal Note**: I and my spouse are fully vaccinated, but we have two children under the age of 12 (including one newborn) who are currently ineligible for the vaccine. While the vaccines are effective at preventing transmission the situation is continuously evolving and  I appreciate everyone's help in minimizing the risks to my family, and any members of your families, who may be more vulnerable. 

---

### Modality Changes

So what happens if we have to revert back to remote learning because of COVID? Ultimately my goal is to structure the course in a way to preserve our flexibility to the fullest extent possible. I have pre-recorded lectures available on my [YouTube channel](https://www.youtube.com/channel/UC7AzCLAufJAMCKkDL77clzw). In the event that we need to move to online only learning students can still view the lectures corresponding to each of the topics we're going to cover this semester. We will also use Zoom where needed to facilitate continued discussion. 

## Required Readings 

Normally I would assign a few books for us to read over the course of the semester, but these are trying times and finances are tight all around. Accordingly the readings for this course will all be available through K-State libraries for free. See the schedule below. 


## Additional Readings

Students should also make an effort to stay informed on current events. Below are a some examples of good sources for keeping up with global events. Please note that some of these publications may be pay-walled, but you should have access through the university library.

  - [BBC News](www.bbc.com/news)
  - [The Monkey Cage](www.washingtonpost.com/news/monkeycage)
  - [Political Violence @ a Glance](www.politicalviolenceataglance.org)
  - [Foreign Affairs](www.foreignaffairs.com)
  - [Foreign Policy Magazine](www.foreignpolicy.com)
  - [The Washington Post](www.washingtonpost.com)
  - [The New York Times](www.nytimes.com)
  - [The Economist](www.economist.com)
  - [The Financial Times](www.ft.com)
  

## Course Requirements

1. _Collaborative Discussion: 30%_

While the course has a heavy lecture component students will also spend time reflecting on class materials and discussing that material with classmates. We will follow a Think-Pair-Share format---Students will have an opportunity to reflect on the course material independently, discuss the material with a classmate, and then communicate a summation of that discussion with the rest of the class. This format should help to reduce the number of contacts individual students have to have for class discussions, and allow us the flexibility to move online if the need arises. 

2. _Content Quizzes: 20%_

Students will complete a series of short quizzes focused on readings and current events. All quizzes are open book and open note assignments and will be completed through Canvas. I will also identify the appropriate readings corresponding to each quiz so you do not have to worry about guessing. 


3. _Photo Blog Journal: 20%_

Students will be required to keep a photo blog. The goal of this assignment is to help students to develop a fuller appreciation for the ways in which their local community connects to the outside world. Here's the gist of the assignment:

  - Identify something around campus, in town, or in the nearby region that connects to US foreign policy and world politics. 
  - Take a picture of it!
  - Write blog post of 800-1000 words in which you:
    - Share the picture
    - Tell us what it is
    - Tell us where it is
    - Explain how it ties into US foreign policy and world politics
    - Share their posts with other students in the class
    
Be creative! There's really no right or wrong here, but it's on you to clearly explain how the subject of your photo ties into the course's subject matter. 
  
To complete this assignment you will have to create a blogger blog and submit the URL to Canvas. We will make sure this content is sharable through Canvas so students can explore each other's work. 
  
4. _The Music of Foreign Policy Wiki: 30%_

The arts are a powerful tool for communicating ideas across space and time, and are often deeply rooted in the politics of the moment. Over the last century music has increasingly come to serve as a vehicle for political speech, both reflecting and driving politics in the modern era. The goal of this project is for students and their groups to build a wiki page on Canvas is dedicated to providing readers with a deeper understanding of the historical, political, and social underpinnings of songs. But there's a catch—since this is a class on US foreign policy, these songs have to be related to US foreign policy in some way. 

In general, each group's wiki page should focus on the following points:

  - *Song/Album*: Identify a song (or if you're feeling ambitious, an entire album) that speaks to some aspect of US foreign policy.
  - *Issue*: Identify the central issue, actors, policy, action, idea, behavior, etc., that is at the core of the song. 
  - *Context*: Once you have identified the central issue, provide some historical background information. This should help readers to better understand the social, political, and economic issues of the time period in question. Who are the key actors? What are the issues or policies of the time that serve as the focus for the song/album? 
  - *Reception*: How was the song/album received at the time of its release? Both critically and with respect to the popular impact that it had at the time?
  - *Legacy*: What is the legacy of the song/album? How does its legacy compare with its impact at the time of its release?


This will serve as your final project. Accordingly, I expect the final content to be detailed, polished, and thoroughly proofed. Your grade will be broken up into components and there will be multiple checkpoints along the way. The schedule is detailed below.

Deadlines: 

- September 10: Song/album choices due (2.5%)
- September 24: Final song/album selection due (2.5%)
- October 29: Annotated bibliography due (10%)
- November 19: Wiki draft due  (5%)
- December 10: Final Wiki page due (10%)
  
  
  
  
## Classroom Policies

1. Be kind. This might seem like an odd thing to list first, but I want to emphasize that this is not a normal class environment---this is a time of unparalleled stress for many of us. We're still in the middle of a global pandemic, many of us are still very much worried about our own health and safety, and many of us are very much worried about the health and safety of our coworkers, friends, and families. Zoom fatigue is also a thing, so please be patient with one another. Let's try to be cognizant of the fact that many of us (if not all) are going to be struggling at various points, and let's work together to make this semester as educational and enjoyable as possible. 

2. Be professional. While we don't have a traditional classroom environment this semester, please try to treat your interactions with everyone as professionally as you would during a normal on-campus experience. This includes not watching TV or playing video games when you are video conferencing with the instructor or your classmates. Everyone's living situation may look a little different this year, and there may be certain things that we can't control about our environments, but let's do what we can to respect one another during our shared time. 

3. Study guides and extra credit. The class is the study guide. I do not offer extra credit.

4. Grade Appeals. If you believe that I have given you an incorrect grade on an exam, you may submit
a written appeal. All appeals must be in writing (they may not be made in person) and must contain
an explanation for why the grade is incorrect. I will also require you to wait 24 hours after receiving a
grade to submit a written appeal. Please note that even if your appeal is granted, I reserve the right
to regrade the assignment in its entirety, meaning that your grade may go up or down depending on
my reevaluation of the assignment.

5. The Syllabus. The syllabus is a living document that can and will be altered throughout the duration
of the course based both on need and design. Generally, this may mean readings will be removed or
added as needed. All changes will be listed on K-State Online, so make sure to check announcements
for any such change. Additionally, I maintain the right to all of my own intellectual property presented
in this course, whether it is the course lectures or this syllabus. Materials from this course ought not
to be reproduced without my expressed permission.

6. Classroom conduct. All student activities in the University, including this course, are governed by
the Student Judicial Conduct Code as outlined in the Student Governing Association By Laws, Article
V, Section 3, number 2. Students who engage in behavior that disrupts the learning environment may
be asked to leave the class.
  
  
## Academic Honesty and Plagiarism
  
All work turned in as a part of this class must be original.

Kansas State University has an Honor & Integrity System based on personal integrity which is presumed to be sufficient assurance in academic matters one’s work is performed honestly and without unauthorized assistance. Undergraduate and graduate students, by registration, acknowledge the jurisdiction of the Honor Integrity System. The policies and procedures of the Honor System apply to all full and parttime
students enrolled in undergraduate and graduate courses on-campus, off-campus, and via distance
learning.

A component vital to the Honor & Integrity System is the inclusion of the Honor Pledge which applies to all
assignments, examinations, or other course work undertaken by students. The Honor Pledge is implied,
whether or not it is stated: “On my honor, as a student, I have neither given nor received unauthorized aid
on this academic work.”

The default in this class is that ALL work will be accomplished individually, UNLESS my permission is given
in advance of an assignment/quiz/exam/take-home exam/final. If you are in doubt, please ask.

A grade of XF can result from a breach of academic honesty. The F indicates failure in the course; the X
indicates the reason is an Honor Pledge violation.

For more information, visit the [Honor & Integrity System home web page](http://www.k-state.edu/honor/).
All work is expected to adhere to the standards outlined by the Kansas State University Handbook. All
written work is subject to online plagiarism checks and will be required for all reaction papers as well as the
final paper. Any cases of plagiarism will receive an automatic failure for the assignment, possible failure in
the course, and will be pursued further through the university for additional punishment.

The Kansas State University Handbook (2008, Appendix F) defines plagiarism as follows:

> Plagiarism is taking credit for someone else’s ideas, work, or words. In a university setting, it
means submitting academic, scholarly, or literary work in which you either claim or imply the
material to be your own, when that is not the case.

The Handbook also provides examples of plagiarism, and steps students can take to avoid it:

[P]lagiarism could include:

  - Buying a term paper, or copying another person’s paper, even if she or he gives you permission.
  - Cutting and pasting information from the internet into your own paper without properly
identifying it as a quotation and properly attributing the source.
  - Using someone else’s concept and presenting it as if it is your own original creativity, without
acknowledging the source of the idea.

While some acts of plagiarism are obvious attempts to deceive, like buying a term paper, others
result from sloppy scholarship or failure to follow proper format for crediting sources. For
example:

  - If you copy directly from another source and acknowledge it in your bibliography or list of
works cited, yet fail to put it in quotation marks, this is plagiarism. The reason is that,
although you have indicated that the idea is someone else’s, the lack of quotation marks
implies that the words are your own.
  - Paraphrasing is summarizing a source so that you have rewritten the material in your own
words but maintained the original author’s ideas. When you paraphrase, you do not need to
put the words in quotation marks, but you do need to properly attribute the original source.
Paraphrasing does not mean simply changing one or two words, or leaving out a sentence,
while the rest remains the same. Even if you identify the original author, but have used faulty
paraphrasing that retains too much of the original, the result is inadvertent plagiarism.
  - If you use another researcher’s specific methodology, you should acknowledge doing so.
While some research protocols are widespread and general enough to need no citation, if
the protocol, analysis, or technique can be attributed to a specific publication it should be
cited.
  - Works of visual art and design, and images of the work, cannot be misrepresented as
original work by another. Such images cannot be used in part or whole without permission
of the artist.

Also note that using papers previously written for another assignment in this, or in any other class, without
the express permission of the professor is grounds for failing an assignment. Using your own material in
this way, and not citing it, is also technically plagiarism and can be treated as such. If you have concerns
regarding any of these issues please do not hesitate to contact me.


## Statement Regarding Wearing of Face Coverings

To protect the health and safety of the K-State community, students, faculty, staff and visitors must wear face coverings over their mouths and noses while on K-State campuses in all hallways, public spaces, classrooms and other common areas of campus buildings, and when in offices or other work spaces or outdoor settings when 6-feet social distancing cannot be maintained. In addition, all students, faculty, and staff are required to take the COVID-19 and Face Mask Safety training. Employees who need reasonable accommodations and assistance related to required face coverings may contact the ADA coordinator at [charlott@k-state.edu](charlott@k-state.edu), and students needing accommodations may contact the Student Access Center at [accesscenter@k-state.edu](accesscenter@k-state.edu).

In classrooms, faculty have the right to deny a student entry into the room if the student is not wearing a face covering. Students not wearing a face covering will be reminded to do so and offered a clean face covering, if one is available. If the student does not comply, the faculty member will ask the student to leave the space, and if available, join the class remotely.  As a last resort, campus police will be called. The faculty members will complete the Code of Conduct form and the Office of Student Life will look further into the issue and take the non-compliance with the request to leave into consideration of further accountability measures. 

At no point should the professor or other students put themselves into an unsafe situation while attempting to enforce the face-covering policy. Manhattan campus police: 785-532-6412



## Citation and Style

All written work is expected to adhere to professional standards. As described above, citing previous work is
key. If you have any questions regarding the appropriate citation style please consult the American Political
Science Association style guide.


## Students with Disabilities

Students with disabilities who need classroom accommodations, access to technology, or information about
emergency building/campus evacuation processes should contact the Student Access Center and/or their
instructor. Services are available to students with a wide range of disabilities including, but not limited to,
physical disabilities, medical conditions, learning disabilities, attention deficit disorder, depression, and anxiety.
If you are a student enrolled in campus/online courses through the Manhattan or Olathe campuses,
contact the Student Access Center at [accesscenter@k-state.edu](accesscenter@k-state.edu), 785-532-6441; for Salina campus, contact
the Academic and Career Advising Center at acac@k-state.edu, 785-826-2649.

If you have any special conditions or needs that will affect your ability to complete the required assignments
outlined herein, please contact the instructor as soon as possible to discuss your needs. I am happy to help
you in any way that I can. With advance notice accommodations can be made prior to deadlines and tests.
Please note, however, that special consideration will not be granted afterwards as a means of addressing a
grade you do not like.

## Contacting the Instructor

The best way to get in touch with me is through email. Please do not hesitate to contact me if you have any questions about course content or your abiliy to complete the assigned material.



## Final Grades

A: 90+

B: 80–89.99

C: 70–79.99

D: 60–69.99

F: 0–59.99


*Tenative course schedule begins on next page*

\clearpage


## Tentative Course Schedule

```{r course calendar, dpi = 400, warning = FALSE, echo = FALSE, message = FALSE, fig.height=9, fig.width=7, fig.pos="center"}

library(ggplot2)
# Create base data frame
dfr <- data.frame(date=seq(as.Date('2021-08-01'), as.Date('2021-12-31'), by = 1))
dfr$day <- factor(strftime(dfr$date,format="%a"), levels=c("Sun","Mon","Tue","Wed","Thu","Fri","Sat"))
dfr$week <- factor(strftime(dfr$date,format="%U"))
dfr$month <- factor(strftime(dfr$date,format="%B"),levels=c("August","September","October", "November", "December"))
dfr$ddate <- factor(strftime(dfr$date,format="%d"))

# Label dates
dfr$comment <- "Available"
dfr$comment[dfr$date < as.Date("2021-08-23") | dfr$date>as.Date("2021-12-10")] <- NA
dfr$comment[dfr$date==as.Date("2021-09-10")] <- "Song/album choice due"
dfr$comment[dfr$date==as.Date("2021-09-24")] <- "Final song/album choice due"
dfr$comment[dfr$date==as.Date("2021-10-29")] <- "Annotated bibliography due"
dfr$comment[dfr$date==as.Date("2021-11-19")] <- "Wiki draft due"
dfr$comment[dfr$date==as.Date("2021-12-10")] <- "Final Wiki due"

# Photo blogs
dfr$comment[dfr$date==as.Date("2021-09-03")] <- "Blog post due"
dfr$comment[dfr$date==as.Date("2021-09-17")] <- "Blog post due"
dfr$comment[dfr$date==as.Date("2021-10-08")] <- "Blog post due"
dfr$comment[dfr$date==as.Date("2021-11-12")] <- "Blog post due"


# Thanksgiving
dfr$comment[dfr$date >= as.Date("2021-11-22") & dfr$date <= as.Date("2021-11-26")] <- "Thanksgiving Break"

dfr$week <- as.numeric(dfr$week) - min(as.numeric(dfr$week) - 1)
dfr$comment <- factor(dfr$comment, levels=c("Blog post due", "Song/album choice due", "Final song/album choice due", "Annotated bibliography due", "Wiki draft due", "Thanksgiving Break", "Final Wiki due"))

ggplot(dfr, aes(x = day, y = reorder(week, -week), fill = comment)) +
  geom_tile(color = "black", size = 0.1) +
  geom_text(aes(label = ddate), size = 3) +
  facet_wrap(. ~ month, scales = "free", ncol = 2) +
  theme_void() +
  theme(legend.position = "bottom",
        legend.direction = "vertical",
        legend.justification = "right",
        axis.text.x = element_text(size = 10, color = "black"),
        axis.text.y = element_text(size = 10, color = "white"),
        strip.text.x = element_text(size = 13, face = "bold", hjust = 0),
        title = element_text(face = "bold")) +
  guides(fill=guide_legend(ncol=2)) +
  scale_fill_brewer(palette = "Set2", na.translate = FALSE) +
  labs(x = "Weekday",
       y = "Week #",
       fill = "Assignment")
```
  
  
  
\clearpage


## `r advdate(mon, 1)`: Course Introduction

1. Watch video on [how to read a social science research paper](https://www.youtube.com/watch?v=pZq6c_shco0)
2. Set up your blog!

## `r advdate(mon, 2)`: Historical background

### Readings and Assignments: 

- Read @fordham2019


## `r advdate(mon, 3)`: The Cold War and US Hegemony

### Readings and Assignments:

- Read @NSC68



## `r advdate(mon, 4)`: The Cold War and Domestic Politics

### Readings and Assignments:

- Read @Fordham2007


## `r advdate(mon, 5)`: The Presidency and Congress

### Readings and Assignments:

- Read @bryan2021
- Read @lupton2017



## `r advdate(mon, 6)`: The Bureaucracy / The State Department

### Readings and Assignments:

- Reading @keane2015

## `r advdate(mon, 7)`: The Defense Department and the Military

### Readings and Assignments:

- Read @brooks2016


## `r advdate(mon, 8)`: The Policymaking Process

### Readings and Assignments:

- Read @marsh2014
- Read @miller2016




## `r advdate(mon, 9)`: The Use of Military Force

### Readings and Assignments:

- Watch [Restrepo](https://watchdocumentaries.com/restrepo/)
- Read @brooks2015


## `r advdate(mon, 10)`: Alliance Relationships

### Readings and Assignments:

1. Read @beckley2015
2. Read [Flynn, Allen, and Martinez Machain (2019)](https://theconversation.com/why-does-the-us-pay-so-much-for-the-defense-of-its-allies-5-questions-answered-127683)


## `r advdate(mon, 11)`: Military Deployments

### Readings and Assignments:

- Read @Allenetal2020
- Read @moon2009


## `r advdate(mon, 12)`: Trade, Immigration, Finance, and Monetary Policy

### Readings and Assignments:

- Read @goldstein2014


## `r advdate(mon, 13)`: Human Rights Issues

### Readings and Assignments:

- Read @Power2001 


## `r advdate(mon, 14)`: Multilateralism and global governance

### Readings and Assignments:

- Read @schultz2017


## `r advdate(mon, 15)`: Thanksgiving Break (No Class)

### Readings and Assignments:

1. Get some rest
2. Enjoy yourselves
3. Eat good food
4. Take a good nap


## `r advdate(mon, 16)`: Foreign Policy in the Age of COVID-19 

### Readings and Assignments:
 
- Read @busby2020 

