---
title: "US Foreign Policy"
subtitle: "Multilateralism in US Foreign Policy"
author: "Michael E. Flynn"
institute: "Kansas State University"
date: "Updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: ["default", "lecture_theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: 16:9
      titleSlideClass: [left, middle, inverse]
---
```{r setup, include = FALSE, echo = FALSE}
library(tidyverse)
library(xaringan)
library(wbstats)
library(tidyquant)
library(plotly)
library(pdftables)
library(pdftools)
library(scales)
library(here)
library(Hmisc)
library(kableExtra)
library(knitr)
library(lubridate)
library(rgdal) # used to read world map data
library(rgeos) 
library(rvest)
library(maptools)
library(ggmap)
library(colorRamps)
library(RColorBrewer)

knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.align="center", echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, dpi=300) 

emissions <- wb(indicator = c("EN.CO2.TRAN.MT", "EN.CO2.MANF.MT", "EN.ATM.CO2E.PC", "NV.AGR.TOTL.ZS"), startdate = 1990, enddate = 2020)

emissions.2 <- emissions %>% 
  filter(indicatorID == "EN.ATM.CO2E.PC") %>% 
filter(country == "High income" | country == "Middle income" | country == "Low income") %>% 
  mutate(country = forcats::fct_relevel(country, "High income", "Middle income", "Low income"))

sysfonts::font_add_google("Oswald", family = "oswald")
showtext::showtext_auto()

basesize <- 12
scale <- 4.5

# Regular plot theme
theme_flynn <- function(){ 
  
      theme_linedraw(base_size = basesize*scale, base_family = "oswald") %+replace% 
        
        theme(plot.title = element_text(face = "bold", size = basesize*scale*1.3, hjust = 0, 
                                        margin = margin(t = 0, b = 0.3, l = 0, r = 0, unit = "cm")),
              plot.subtitle = element_text(size = basesize*scale, hjust = 0, 
                                           margin = margin(t = 0, b = 0.3, l = 0, r = 0, unit = "cm")),
              plot.caption = element_text(face = "italic", 
                                          size = basesize*scale*0.6, 
                                          hjust = 1, lineheight = 0.35, 
                                          margin = margin(t = 0.2, b = 0, l = 0, r = 0, unit = "cm")),
              strip.background = element_rect(fill = "gray80", color = "black", size = 0.3),
              strip.text = element_text(color = "black", face = "bold", hjust = 0, margin = margin(t = 0.2, b = 0.2, l = 0.2, r = 0.2, unit = "cm")),
              panel.grid.major = element_line(color = "gray70", size = 0.15),
              panel.grid.minor = element_line(color = "gray90", size = 0.1),
              axis.title = element_text(face = "bold", size = basesize*scale),
              axis.title.y = element_text(angle = 90, margin = margin(t = 0, r = 0.5, b = 0, l = 0, unit = "cm")),
              axis.title.x = element_text(margin = margin(t = 0.5, r = 0, b = 0, l = 0, unit = "cm")),
              axis.ticks = element_line(size = 0.2),
              axis.ticks.length = unit(0.2, "cm"),
              panel.border = element_rect(fill = NA, size = 0.3),
              plot.margin = margin(0.5, 0.5, 0.5, 0.5, unit = "cm"),
              legend.title = element_text(face = "bold", hjust = 0, margin = margin(t = 0, b = -0.5, l = 0, r = 0, unit = "cm")),
              legend.margin=margin(t=0, b=0, r=0, l=0),
              legend.box.margin=margin(t=-10, b=-10, l=-10, r=-10))
  }


```


# Lecture Overview

1. Defining multilateralism

2. Historical review of multilateralism in US foreign policy

3. Costs and benefits

4. Why Multilateralism?



---
class: center, middle, inverse

# Defining Multilateralism



---
class: left, top

# Defining Multilateralism

> **Multilateralism** refers to efforts to coordinate behavior between three or more actors.



---
class: center, middle, inverse

# Historical Overview





---
class: left, top

# Historical Overview

Multilateralism in US foreign policy

- For the most part, US foreign policy is mostly characterized by unilateralism through World War I

- World War I marks a slight turn towards multilateralism, but there's a lot of resistance

- After World War II multilateralism becomes a central component of US foreign policy




---
class: left, top

# Historical Overview

Multilateralism in security, diplomatic, and economic affairs

- North Atlantic Treaty Organization (NATO)

- United Nations (UN)

- General Agreement on Tariffs and Trade (GATT)

- International Monetary Fund (IMF)

- World Bank




---
class: center, middle, inverse

# Costs and Benefits




---
class: top, left

# Costs and Benefits

Multilateralism has a range of shorter-term benefits

- Pooling resources

- Distribute costs

- Enhanced legitimacy

--

But there are also costs

- Decreased autonomy/constraints on behavior

- Possible costs via side payments 

- More diffuse policy benefits


---
class: center, middle

```{r}

fat.data <- readxl::read_xlsx(here("../../../Data Files/icasualties/afghan-iraq-fatalities.xlsx")) %>%
  mutate(NonUS = Total - US) %>% 
  select(Year, War, US, NonUS) %>% 
  pivot_longer(cols = c(3,4),
               names_to = "country",
               values_to = "deaths")

ggplot(fat.data, aes(x = Year, y = deaths, fill = country)) +
  geom_bar(stat = "identity", position = "stack", color = "black", size = 0.1) +
  facet_wrap(. ~ War) +
  theme_flynn() +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Year",
       y = "Count",
       fill = "Country",
       title = "Coalition deaths in Afghanistan and Iraq, 2001-2020",
       caption = "Data obtained from icasualties.org")


```


???

# Notes

- Example of burden sharing in AFghanistan and comparison case of Iraq
- In Afghanistan US endured majority of casualties, but coalition partners also took on a sizeable share of casualties
- In Iraq the US went in with relatively few partner states dedicated to combat operations
- Note this doesn't include other areas of burden sharing like financial support





---
class: center, middle

```{r coalition fatalities percent}

fat.perc <- readxl::read_xlsx(here("../../../Data Files/icasualties/afghan-iraq-fatalities.xlsx")) %>%
  mutate(percent = US/Total)

ggplot(fat.perc, aes(x = Year, y = percent, color = War, linetype = War)) +
  geom_line(size = 1.5) +
  theme_flynn() +
  scale_color_brewer(palette = "Set2") +
  scale_y_continuous(labels = percent_format()) +
  guides(linetype = FALSE) +
  labs(x = "Year",
       y = "Percent",
       color = "Conflict",
       title = "US fatalities as a percentage of total coalition fatalities",
       subtitle = "Coalition warfighting can spread risk to many states.",
       caption = "Data obtained from icasualties.org")
```


???

# Notes

- Another example from Iraq and Afghanistan
- US fatalities consistently high as a percentage of total fatalities in Iraq
- In Afghanistan allied states suffered a much higher proportion of the total casualties over time, particularly during the "surge" years around 2009 and 2010.




---
class: top, left

# Costs and Benefits

But even short-term tradeoffs can have long-term benefits

The US was the dominant military and economic power following the War, but how best to wield that power?

- A retreat from engagement to unilateralism like after World War I?
  
- Assert dominant power to control other countries?
  
- Something else?

>**Liberal Hegemony**: A term coined by John Ikenberry to describe the nature of the international order that obtained following World War II where the US acted as the dominant power presiding over a rules-based international system to which it subordinated itself (at least in part).



---
class: left, top

# Costs and Benefits

Multilateralism was part of a strategy to balance and advance multiple objectives

- Secure Western allies

- Deter Soviet aggression

- Perpetuate the United States' dominant position for the long-term

Liberal Hegemony entailed a few key components

- US sacrificed some autonomy, binding itself to a set of rules-based arrangements in different policy areas

- By binding itself the US obtained the buy-in from other countries, increasing the legitimacy and long-term viability of US-led projects




---
class: center, middle, inverse

# Why Multilateralism?



---
class: left, top

# Why Multilateralism?

Lots of different and important issues, but what do they have in common?

--

Well, there's a shared, fundamental problem underlying these issues

--

Collective action and coordination problems




---
class: left, top

# Basic problems



>**Collective Action Problems:** Interactions where individuals/groups would all benefit from cooperating to produce some good, but where individual players have an incentive to defect.


>**Coordination Problems**: Coordination therefore, is the orderly arrangement of group effort, to provide unity of action in the pursuit of a common purpose. (Mooney 1947)

> ALSO

> [A] coordination problem exists if achievement of a particular organizational goal requires that each actor select the appropriate action, and the goal is not fully achieved if all members of the group do not select goal-fulfilling actions. (Camerer and Knez 1997)





---
class: center, middle


```{r}

pd.table <-data.frame(`Cooperate` = c("4,4", "5,1"),
                      `Defect` = c("1,5", "2,2"))

rownames(pd.table) <- c("Cooperate", "Defect")

kable(pd.table, caption = "Prisoner's Dilemma Example", booktabs = TRUE, "html",
               align = c("c", "c"),
      escape = FALSE) %>% 
  kable_styling() %>% 
  row_spec(0, bold = TRUE, color = "white", background = "#512888") %>% 
  column_spec(1, bold = TRUE, color = "white", background = "#512888") %>% 
  column_spec(2, width = "3cm") %>% 
  column_spec(3, width = "3cm")

```



---
class: center, middle

```{r}

# For some reason the escale=FALSE line is required below to get the cell_spec coloring to work.

pd.table <-data.frame(`Cooperate` = c("4,4", "5,1"),
                      `Defect` = c("1,5", "2,2"))  %>% 
  mutate(`Cooperate` = cell_spec(`Cooperate`, "html", color = ifelse(`Cooperate` == "5,1", "red", "black")))

rownames(pd.table) <- c("Cooperate", "Defect")


kable(pd.table, caption = "Prisoner's Dilemma Example", booktabs = TRUE, "html",
               align = c("c", "c"),
      escape = FALSE) %>% 
  kable_styling() %>% 
  row_spec(0, bold = TRUE, color = "white", background = "#512888") %>% 
  column_spec(1, bold = TRUE, color = "white", background = "#512888") %>% 
  column_spec(2, width = "3cm") %>% 
  column_spec(3, width = "3cm") 

```



---
class: center, middle

```{r}

# For some reason the escale=FALSE line is required below to get the cell_spec coloring to work.

pd.table <-data.frame(`Cooperate` = c("4,4", "5,1"),
                      `Defect` = c("1,5", "2,2"))  %>% 
  mutate(`Defect` = cell_spec(`Defect`, "html", color = ifelse(`Defect` == "1,5", "red", "black")))

rownames(pd.table) <- c("Cooperate", "Defect")


kable(pd.table, caption = "Prisoner's Dilemma Example", booktabs = TRUE, "html",
               align = c("c", "c"),
      escape = FALSE) %>% 
  kable_styling() %>% 
  row_spec(0, bold = TRUE, color = "white", background = "#512888") %>% 
  column_spec(1, bold = TRUE, color = "white", background = "#512888") %>% 
  column_spec(2, width = "3cm") %>% 
  column_spec(3, width = "3cm") 

```




---
class: center, middle

```{r}

# For some reason the escale=FALSE line is required below to get the cell_spec coloring to work.

pd.table <-data.frame(`Cooperate` = c("4,4", "5,1"),
                      `Defect` = c("1,5", "2,2")) %>% 
  mutate(`Defect` = cell_spec(`Defect`, "html", color = ifelse(`Defect` == "2,2", "red", "black")))

rownames(pd.table) <- c("Cooperate", "Defect")


kable(pd.table, caption = "Prisoner's Dilemma Example", booktabs = TRUE, "html",
               align = c("c", "c"),
      escape = FALSE) %>% 
  kable_styling() %>% 
  row_spec(0, bold = TRUE, color = "white", background = "#512888") %>% 
  column_spec(1, bold = TRUE, color = "white", background = "#512888") %>% 
  column_spec(2, width = "3cm") %>% 
  column_spec(3, width = "3cm") 

```





---
class: left, top

# Why Multilateralism?

1. Responding to adversaries

2. Trade and investment policy

3. Peacekeeping

4. Human rights and trafficking

5. Global climate change







---
class: center, middle

```{r}

data.temp.anom <- read_csv(here("../../../Data Files/NOAA Data/1880-2016.csv"), skip = 3) %>% 
  mutate(group = ifelse(Value > 0, "Positive", "Negative"))

data.temp <- readxl::read_xls(here("../../../Data Files/US Government Files/EPA/vostok.1999.temp.dat.xls"), skip = 0)
colnames(data.temp)[2] <- "time"
colnames(data.temp)[4] <- "temp"
data.temp <- data.temp %>% 
  mutate(time = as.numeric(0 - time),
         value = as.numeric(temp),
         group = as.factor("Temp")) %>% 
  subset(select = c("time", "value", "group"))


# Time variable had to be converted from factor to numeric
# Have to use paste function
data.ice <- read_csv(here("../../../Data Files/US Government Files/EPA/ghg-concentrations_fig-1.csv"), skip = 7)
colnames(data.ice)[1] <- "time"
data.ice <- data.ice %>% 
  mutate(co2 = apply(., 1, max, na.rm = TRUE),
         time = as.numeric(paste(time)),
         value = as.numeric(co2),
         group = as.factor("CO2")) %>% 
  subset(select = c("time", "group", "value"))

data.combined <- rbind(data.ice, data.temp)


# The second scale was tricky. You have to reverse what you do for the rescaling in the geom_line section
# Also note that, since you multiply by 10 BEFORE adding 250, you have to divide by 10 and subtract 25
ggplot() +
  geom_line(data = subset(data.combined, group == "CO2"), aes(x = time, y = value, group = group, colour = group)) +
  geom_line(data = subset(data.combined, group == "Temp"), aes(x = time, y = ((value * 10) + 250), group = group, colour = group), alpha = .6) +
  geom_hline(yintercept = 250) +
  scale_y_continuous(breaks = seq(150, 400, 50), limits = c(150, 400), sec.axis = sec_axis(~./10 - 25, name = "Temperature Anomaly (°C)", breaks = seq(-10, 16, 2))) +
  scale_x_continuous(breaks = seq(-800000, 0, 100000), labels = scales::comma) +
  theme_flynn() +
  theme(legend.position = "bottom",
        legend.margin = margin(t = -0.5, b = 0.5, unit = "cm"),
        panel.grid.major = element_line(colour = "gray70", size = .1),
        panel.grid.minor = element_line(colour = "gray80", size = .1)) +
  scale_color_brewer(palette = "Set1") +
  labs(colour = "Measure",
       title = "Global temperature anamoly and atmospheric CO2 levels",
       subtitle = "CO2 levels and temperature anomalies track one another closely over the last 400,000 years",
       y = "CO2 (ppm)",
       x = "Time",
       caption = "Ice core temperature data obtained from:
       http://cdiac.ess-dive.lbl.gov/ftp/trends/temp/vostok/vostok.1999.temp.dat
       CO2 data obtained from: https://www.epa.gov/sites/production/files/2016-08/ghg-concentrations_fig-1.csv
       Note: Temperature data ends 2000 years prior to present day.") 




```




---
class: center, middle

```{r CO2 emissions, warning=FALSE, echo = FALSE, error=FALSE}


ggplot(data = emissions.2) +
  geom_area(aes(x = as.numeric(date), y = value, group = country, fill = country), colour = "black", size = 0.1) +
  theme_flynn() +
  labs(x = "Year",
       y = "Metric tons per capita",
       fill = "Group",
       title = "CO2 emissions by income group, 1990-2016",
       subtitle = "How does the distribution of wealth and emissions affect responsibility to reduce emissions?",
       caption = "Data obtained from World Bank data series EN.ATM.CO2E.PC") +
  scale_fill_brewer(palette = "Spectral") +
  scale_x_continuous(breaks = seq(1990,2020,5))


```





---
class: center, middle

```{r Adirondack suflate levels}

# Note I had to rename a complicated column here. Basic rename didn't work, had to use col number.
data <- read_csv(here("../../../Data Files/Adirondack Lake Survey/2016-1992_ALTM_WEB.csv"), skip = 11)
colnames(data)[6] <- "Sample Date"
colnames(data)[8] <- "Sulfate"
colnames(data)[4] <- "Station"
colnames(data)[5] <- "Depth"

# The date conversation is tricky here
# The format part has to match the existing format from the imported data (here it's a ".")
# The sep line tells it what to convert the separator to
data <- data %>% 
  subset(`Station` == 1) %>% 
  mutate(`Sample Date` = as.Date(`Sample Date`, format = "%Y.%m.%d", sep = "-", origin = "1970-01-01"))

data.1 <- data %>% 
  subset(select = c("Sample Date", "Sulfate"))

data.2 <- read_csv(here("../../../Data Files/Adirondack Lake Survey/ALS_SUMMER_SURFACE_CHEM_1984-87.csv"), skip = 3)
colnames(data.2)[3] <- "Sample Date"
colnames(data.2)[6] <- "Sulfate"
colnames(data.2)[4] <- "long"
colnames(data.2)[5] <- "lat"
data.2 <- data.2 %>% 
  subset(select = c("Sample Date", "Sulfate")) %>% 
  mutate(`Sample Date` = as.Date(`Sample Date`, format = "%m/%d/%y", sep = "-", origin = "1970-01-01"))

data.combined <- rbind(data.1, data.2)


# Combined Point data
ggplot(data = data.combined, aes(x = `Sample Date`, y = Sulfate)) +
  geom_point(alpha = .2) +
  geom_smooth(method = "lm", formula = y ~ splines::bs(x, 3), se = FALSE) +
  geom_vline(xintercept = as.Date("1990-11-15"), colour = "red") +
  annotate("text", x = as.Date("1992-01-01"), y = 16, label = "1990 Clean Air Act \nammendments signed", hjust = 0, size = basesize, lineheight = 0.4) +
  theme_flynn() +
  labs(title = "Sulfate levels in Adirondack lakes, 1984-2016",
       subtitle = "Public policy can help to mitigate or even reverse problems resulting from pollution",
       y = "Sulfate Concentration (mg/L)",
       caption = "Data obtained from the Adirondack Lakes Survey Corporation: \n http://www.adirondacklakessurvey.org/data/complete/1992-2016_ALTM_Chemistry.zip \nhttp://www.adirondacklakessurvey.org/data/sumsurf/sumsurf_chem/ALS_Summer_surface_chem_1984-87.zip \nGraph depicts surface sample values only.") +
  scale_x_date(breaks = seq(as.Date("1984-01-01"), as.Date("2017-01-01"), by = "4 years"), 
                            labels = date_format("%b %Y"),
               limits = c(as.Date("1984-01-01"), as.Date("2017-01-01")))


```
